---
title: "Lab 04 - Data Exploration"
author: Yongpeng Fu
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: simplex
    number_sections: false
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## 0 load the package 
```{r}
library(tidyverse)
```

## 1 create a tibble 

```{r}
# load the data 
df = read_csv("lab-04-firm.csv")
# first six rows 
head(df)
# column names
colnames(df)
```

## 2 calculate summary statistics 

```{r}
# 2.1 average net income   
mean(df$NI)
# 2.2 first decile of net income  
quantile(df$NI,0.1)
# 2.3 summary of net income    
summary(df$NI)
```

## 3 compute counts and proportion of logical values    
```{r}
# 3.1 number of rows 
sum(df$DV > 0 & df$NI > 10000)
# 3.2 proportion of rows 
mean(df$DV > 0 & df$NI > 10000)
```

## 4 summarize data with dpylr  

### 4.1 average and maximum net income 
```{r}
summarize(df,
          NI_mean = mean(NI),
          NI_max = max(NI),
          n_rows = nrow(df),
          n_firms = n_distinct(TIC))
```

### 4.2 average and minimum net income by firm 
```{r}
df_by_TIC = df %>%
  group_by(TIC) %>%
  summarize(NI_mean = mean(NI),
            NI_min = min(NI),
            n_rows = n())
df_by_TIC
```

## 5 combine multiple operations with the pipe 

### 5.1 average net income by firm  
```{r}
df %>%
  group_by(TIC) %>%
  summarise(NI_mean = mean(NI)) %>%
  arrange(desc(NI_mean))

```

### 5.2 most recent return on assets by firm 
```{r}
df %>%
  mutate(ROA = NI/AT) %>%
  group_by(TIC) %>%
  arrange(desc(FYEAR),.by_group = TRUE) %>%
  summarise(recent_ROA = first(ROA))
  

```

## 6 combine firm fundamentals with executives’ compensation	

### 6.1 load executive compensation data 
```{r}
# load the data 
manager = read_csv("manager.csv")
# first six rows 
head(manager)
# column names
names(manager)
```

### 6.2  frequency counts of firm
```{r}
table(df$TIC)

table(manager$TICKER)

```

### 6.3 frequency counts of year
```{r}
table(df$FYEAR)

table(manager$YEAR)

```

### 6.4 inner join 
```{r}
# inner join 
merged = inner_join(manager, df,
                     by = c("TICKER" = "TIC",
                            "YEAR"   = "FYEAR"))
# first six rows 
head(merged)
# column names 
names(merged)
```

### 6.5 average salary and return on assets during 2017 and 2018 
```{r}
merged %>%
  filter(YEAR == 2017 | YEAR == 2018) %>%
  mutate(ROA = NI/AT) %>%
  group_by(TICKER) %>%
  summarise(SALARY_mean = mean(SALARY),
            ROA_mean = mean(ROA),
            n_rows = n()) %>%
  arrange(desc(ROA_mean),.by_group = TRUE)

```


