---
title: "Lab 09 - Logistic Regression"
author: Yongpeng Fu
output:
  html_document:
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## 0 load the package and suppress scientific notation 
```{r}
# load packages 
library(caret,gains)
# not display scientific notation
options(scipen=999)
```

## 1 create a data frame
```{r}
# load the data
# first six rows
ebay.df = read.csv("eBayAuctions.csv",stringsAsFactors = FALSE)
head(ebay.df)
# variable names 
names(ebay.df)

summary(ebay.df)
```

## 2 apply a function to each group 
```{r}
# mean of the binary outcome for each day of the week of auction close  
#tapply(binary outcome variable, grouping variable, function)
tapply(ebay.df$Competitive., ebay.df$endDay, mean) 

# mean of the binary outcome for each auction category  
tapply(ebay.df$Competitive., ebay.df$Category, mean)

```

## 3 reduce levels in the day of week that the auction closed  
```{r}
# combine Sunday and Friday into a single category called Sun_Fri
ebay.df$endDay[ebay.df$endDay == "Sun"] <- "Sun_Fri"
ebay.df$endDay[ebay.df$endDay == "Fri"] <- "Sun_Fri"
```

## 4 reduce levels in auction categories 
```{r}
# combine Business/Industrial and Computers into a single category called Computers

# unique(ebay.df$Category)

ebay.df$Category[ebay.df$Category == "Business/Industrial"] <- "Computer"

# combine Antique/Art/Craft and Collectibles into a single category called collectibles

ebay.df$Category[ebay.df$Category == "Antique/Art/Craft"] <- "Collectibles"
```

## 5 convert Duration to a categorical or factor variable 
```{r}
# convert Duration from a numeric variable to a categorical or factor variable  
ebay.df$Duration <- as.factor(ebay.df$Duration)
# variable types 
str(ebay.df$Duration)
```

## 6 data partition
```{r}
# set seed for reproducing the partition 
set.seed(1)
# takes a sample of row numbers for the training set 
n <- nrow(ebay.df)
train.index <- sample(1:n, 0.6 * n)
# training set 
train.df <- ebay.df[train.index, ]
# test set 
test.df  <- ebay.df[-train.index, ]
```

## 7 fit a logistic regression model 
```{r,warning=FALSE}
# logistic regression using all the variables
reg <- glm(Competitive. ~ ., data = train.df, family = "binomial")
# summary table 
summary(reg)
```

## 8 generate predicted probabilities for records in the test set 
```{r}
# predicted probabilities 
pred <- predict(reg, newdata = test.df, type = "response")
# first six values 
head(pred)
```

## 9 confusion matrix 
```{r}
pred.class <- ifelse(pred > 0.5, 1, 0)

confusionMatrix(
  data = as.factor(pred.class),                # 预测结果
  reference = as.factor(test.df$Competitive.), # 实际结果
  positive = "1"                               # 把 "1" 当成正类
)
```

## 10 creating a gain table [optional]
```{r}
library(gains)
# gain table 
gain <- gains(actual = test.df$Competitive.,
              predicted = pred,
              groups = 10)
# cumulative percentage of competitive auctions 
gain$cume.pct.of.total
# cumulative number of auctions 
gain$cume.obs
```

## 11 plot a lift chart [optional]
```{r}
# plot the cumulative number of competitive auctions against the cumulative number of auctions

yvals <- c(0, gain$cume.pct.of.total * sum(test.df$Competitive.))

xvals <- c(0, gain$cume.obs)

plot(xvals, yvals, type="l",
     xlab="Cumulative number of auctions",
     ylab="Cumulative number of competitive auctions")

# add a baseline curve 
lines(c(0, dim(test.df)[1]),
      c(0, sum(test.df$Competitive.)))

```